A Way Forward in Parallelising Dynamic Languages

### 2014-05-05 talk/icooolps2014/position-paper.tex
Tweaks tweaks tweaks

  Here we want to compare different approaches to replacing the GIL
  with a technology that allows parallel execution. We look at
  fine-grained locking, shared-nothing, and transactional memory (TM)
  approaches. We argue that software-based TM systems are the most
  promising, especially since they also enable the introduction of
  atomic blocks as a better synchronisation mechanism in the language.
\end{abstract}

@@

The approach that wins in the end should perform similarly for
single-threaded execution as compared to the GIL and be able to
execute code in parallel on multiple cores. Furthermore, we will also
take into account the compatibility to existing code that already uses
threads for concurrency, as well as the changes that are required to
the interpreter itself.

These requirements are not easy to meet. We argue that STM is the
overall winner. While it has a big performance problem currently, it
gets more points in all the other categories. We think that it is the
only solution that also provides a better synchronisation mechanism to
the application in the form of paralleliseable atomic blocks.

@@

\begin{table*}[ht]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    & \textbf{GIL} & \textbf{Fine-grained locking}
    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
    \hline
    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
    \hline
    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
    \hline
    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
    \hline
    Better synchronisation        & o    & o  & o    & o  & ++   \\
    \hline
    Implementation                & ++   & -  & ++   & ++ & ++   \\
    \hline
    External libraries            & ++   & ++ & ++   & ++ & ++   \\
    \hline
  \end{tabular}
  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
  \label{tab:comparison}
\end{table*}

@@

Jython~\cite{webjython} is one project that implements an
interpreter for Python on the JVM\footnote{Java Virtual Machine} and
that uses fine-grained locking to correctly synchronise the
interpreter. For a language like Python, one needs quite a few,
carefully placed locks. Since there is no central location, the
complexity of the implementation is quite a bit greater compared to
using a GIL. Integrating external, non-thread-safe libraries should
however be very simple too. One could simply use one lock per library
to avoid this issue.

@@

There are also approaches that work around the GIL instead of trying
to replace it. If an application can be split into completely
independent parts that only very rarely need to share anything, or
only do so via an external program like a database, then it is
sensible to have one GIL per independent part. As an example of such
an approach we look at the
\emph{multiprocessing}\footnote{https://docs.python.org/2/library/multiprocessing.html}
module of Python. In essence, it uses process-forking to provide the
application with multiple interpreters that can run in parallel.
Communication is then done explicitly through pipes.

Obviously not every application fits well into this model and its
applicability is therefore quite limited. Performance is good as
long as the application does not need to communicate a lot, because
inter-process communication is relatively expensive. Also the
implementation of this approach is very cheap since one can
actually take an unmodified GIL-supported interpreter and run
multiple of them in parallel. That way, we also inherit the
easy integration of external libraries without any changes.
While the model of explicit communication is often seen as a
superior way to synchronise concurrent applications because
of its explicitness, it does not actually introduce a better
synchronisation mechanism for applications.

@@

The performance of HTM is pretty good as it does not introduce much
overhead ($<40\%$ overhead~\cite{odaira14}). And it can transparently
parallelise existing applications to some degree. The implementation
is very straight-forward because it directly replaces the GIL in a
central place. HTM is also directly compatible with any external
library that needs to be integrated and synchronised for use in
multiple threads. The one thing that is missing is support for a
better synchronisation mechanism for the application. It is not
possible in general to expose the hardware-transactions to the
application in the form of atomic blocks, because that would require
much longer transactions.

@@

\paragraph{STM} provides all the same benefits as HTM except for its
performance.  It is not unusual for the overhead introduced by STM to
be between 100\% to even 1000\% ~\cite{cascaval08,drago11}. While STM
systems often scale very well to a big number of threads and
eventually overtake the single-threaded execution, they often provide
no benefits at all for low numbers of threads (1-8). There are some
attempts ~\cite{warmhoff13,spear09} that can reduce the overhead a lot,
but scale badly or only for certain workloads. Often the benefits
on more than one thread are too little in real world applications.

@@

While one can argue that STM requires the insertion of read and write
barriers in the whole program, this can be done automatically and
locally by a program transformation~\cite{felber07}. There are attempts
to do the same for fine-grained locking~\cite{bill06} but they require
a whole program analysis since locks are inherently non-composable.
The effectiveness of these approaches is doubtful in our use case,
since we execute bytecode instructions in any order defined by a
script only known at runtime. This makes it close to impossible to
order locks consistently or to know in advance which locks a
transaction will need.

@@

One way to get more performance is to develop STM systems that make
better use of low-level features in existing OS kernels.  We are
currently working on a STM system that makes use of several such
features like virtual memory and memory segmentation.  We further
tailor the system to the discussed use case which gives us an
advantage over other STM systems that are more general. With this
approach, initial results suggest that we can keep the overhead of STM
below 50\%. A hybrid TM system, which also uses HTM to accelerate
certain tasks, looks like a very promising direction of research
too. We believe that further work to reduce the overhead of STM is
very worthwhile.

@@

>>>>

  Here we want to compare different approaches to replacing the GIL
  with a technology that allows parallel execution. We look at
  fine-grained locking, shared-nothing, and transactional memory (TM)
  approaches. We argue that software-based TM systems are the most
  promising, especially since they also enable the introduction of
  large, parallelisable atomic blocks as a better synchronisation
  mechanism in the language.
\end{abstract}

@@

The approach that wins in the end should perform similarly for
single-threaded execution as compared to the GIL and be able to
execute code in parallel on multiple cores. Furthermore, we will also
take into account the compatibility with existing code that may already use
threads for concurrency, as well as the changes that are required to
the interpreter itself.

These requirements are not easy to meet. We argue that STM is the
overall winner. While it has a big performance problem currently, it
gets more points in all the other categories. We think that it is the
only solution that also provides a better synchronisation mechanism to
the application in the form of parallelisable atomic blocks.

@@

\begin{table*}[ht]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    & \textbf{GIL} & \textbf{Fine-grained locking}
    & \textbf{Shared-nothing} & \textbf{HTM} & \textbf{STM}\\
    \hline
    Performance (single threaded) & ++   & +  & ++   & ++ & -{-} \\
    \hline
    Performance (multithreaded)   & -{-} & +  & +    & +  & +    \\
    \hline
    Existing applications         & ++   & ++ & -{-} & ++ & ++   \\
    \hline
    Better synchronisation        & o    & o  &      & o  & ++   \\
    \hline
    Implementation                & ++   & -  & ++   & ++ & ++   \\
    \hline
    External libraries            & ++   & ++ & ++   & ++ & ++   \\
    \hline
  \end{tabular}
  \caption{Comparison between the approaches (-{-}/-/o/+/++)}
  \label{tab:comparison}
\end{table*}

@@

Jython~\cite{webjython} is one project that implements an
interpreter for Python on the Java Virtual Machine (JVM) and
that uses fine-grained locking\footnote{The performance impact of
fine-grained locking is milder in Java than it would be in a typical piece
of C code; see e.g.~\cite{biased}.} to correctly synchronise the
interpreter. For a language like Python, one needs quite a few,
carefully placed locks. Since there is no central location, the
complexity of the implementation is quite a bit larger compared to
using a GIL. Integrating external, non-thread-safe libraries should
however be very simple too. One could simply use one lock per library
to avoid this issue.

@@

There are also approaches that work around the GIL instead of trying
to replace it. If an application can be split into completely
independent parts that only very rarely need to share anything, or
only do so via an external program like a database, then it is
sensible to have one GIL per independent part. At the extreme, there
are applications that parallelise perfectly simply by running
independent processes; some web servers and some numeric computations
do. We will consider here a slightly more general approach: the
\emph{multiprocessing}\footnote{https://docs.python.org/2/library/multiprocessing.html}
module of Python. In essence, it uses process-forking to provide the
application with multiple interpreters that can run in parallel.
Communication is then done explicitly through pipes.

The model of explicit communication is sometimes seen as a superior
way to synchronise concurrent applications because of its explicitness.
However, not every application fits well into this model and its
applicability is therefore limited. Performance is good as
long as the application does not need to communicate a lot, because
inter-process communication is relatively expensive. Also the
implementation of this approach is very cheap since one can
actually take an unmodified GIL-supported interpreter and run
several of them in parallel. That way, we also inherit the
easy integration of external libraries without any changes.

@@

The performance of HTM is pretty good as it does not introduce much
overhead ($<40\%$ overhead~\cite{odaira14}). And it can transparently
parallelise existing applications to some degree. The implementation
is very straight-forward because it directly replaces the GIL in a
central place. HTM is also directly compatible with any external
library that needs to be integrated and synchronised for use in
multiple threads. The one thing that is missing is support for a
better synchronisation mechanism for the application. It is not
reasonable in general to expose the hardware-transactions to the
application in the form of atomic blocks, because doing so would
require the system to support much longer transactions.

@@

\paragraph{STM} provides all the same benefits as HTM except for its
performance.  It is not unusual for the overhead introduced by STM to
be between 100\% to even 1000\% ~\cite{cascaval08,drago11}. While STM
systems often scale very well to a big number of threads and
eventually overtake the single-threaded execution, they often provide
no benefits at all for low numbers of threads (1-8). There are some
attempts ~\cite{warmhoff13,spear09} that can reduce the overhead a lot,
but scale badly or only for certain workloads. Often the benefits
on more than one thread are too small in real world applications.

@@

On the implementation level,
while one can argue that STM requires the insertion of read and write
barriers in the whole interpreter, this can be done automatically and
locally by a program transformation~\cite{felber07}. There are attempts
to do the same for fine-grained locking~\cite{bill06} but they require
a whole program analysis since locks are inherently non-composable
--- and their effectiveness is doubtful in our use case,
since we execute bytecode instructions in any order defined by a
script only known at runtime. This makes it close to impossible to
order locks consistently or to know in advance which locks a
transaction will need.

@@

One way to get more performance is to develop STM systems that make
better use of low-level features in existing OS kernels.  We are
currently working on a STM system that makes use of several such
features like virtual memory and memory segmentation.  We further
tailor the system to the discussed use case which gives us an
advantage over other STM systems that are more general. With this
approach, initial results suggest that we can keep the overhead of STM
well below 50\%. A hybrid TM system, which also uses HTM to accelerate
certain tasks, looks like a very promising direction of research
too. We believe that further work to reduce the overhead of STM is
very worthwhile.

@@

\bibitem{biased}
  Kenneth Russell and David Detlefs. 2006. Eliminating
  synchronization-related atomic operations with biased locking and
  bulk rebiasing. \emph{In Proceedings of the 21st annual ACM SIGPLAN
    conference on Object-oriented programing, systems, languages, and
    applications} (OOPSLA '06).
